---
title: "prereg analysis"
author: "Peter Steiglechner"
date: "19.05.2026"
engine: knitr
format:
  html:
    code-fold: true
---

```{r}
#| output: false


load_or_install <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
pkgs <- c(
  "tidyverse",
  "lme4",
  "lmerTest",
  "nloptr",
  "httpgd",
  "languageserver"
)

invisible(lapply(pkgs, load_or_install))

options(vsc.viewer = TRUE)

```

Configuration
```{r}

questions_sc = c(
    "climate_concern",
    "gay_marriage",
    "rights_indep_integration",
    "econ_inequality",
    "regulate_internet",
    "east_germans"
    )
delta_cols <- paste0("deltaX_", questions_sc)

n_issues <- length(delta_cols)

parties = c("Left Party", "BSW", "Green Party", "SPD", "FDP", "CDU/CSU", "AfD", "No party", "Other party", "Refuse to say/No answer")
partiesVars = c("LeftParty", "BSW", "GreenParty", "SPD", "FDP", "CDU/CSU", "AfD")

```

# Load Data
```{r}
df_partic <- read.csv("processed_data/2026-05-13_data_processed_participant.csv")

inds_bothwaves <- df_partic %>%
  count(id) %>%
  filter(n == 2) %>%
  pull(id)

df_diff <- read.csv("processed_data/2026-05-13_data_processed_differences.csv")

# View(df_diff)
df_partic |> nrow()
df_partic$id |> unique() |> length()
df_diff$id |> unique() |> length()
```

# Analysis Theory

\section{Inferred variables}

The key variables we are interested in are how much weight participants implicitly assign to belief differences on each of the six issues when placing individuals on the political map. We infer these weights, $\alpha_{n,q}$, from the participant $n$'s responses by comparing the map distances, $d_{n, ij}$, of individuals $i$ and $j$ with their belief differences, $\lvert x_{n,i,q}-x_{n,j,q}\rvert$, on all issues $q$ as reported by the participant. In particular, we model the map distances between dots $i$ and $j$ for participant $n$ as a function of their belief differences on issues $q$ 
\begin{equation} \label{eq:distance}
    d_{n, ij} = f_n\left( \, g_n \right(\, \lvert x_{n,i,q} - x_{n,j,q} \, \rvert \left)\, \right) \ .
\end{equation}

Participant-level function $g_n$ aggregates the belief differences on the six items $q$ between individuals $i$ and $j$ into a (normalised) `overall belief distance', that is, $g_n: [0,1]^6 \mapsto [0,1]$. 
Here, we assume $g_n$ as weighted $L1$-norm (or weighted Manhattan/city-block distance):
\begin{equation} \label{eq:weighted}
    g_n (\lvert x_{n,i,q} - x_{n,j,q}\rvert) = \sum_q \alpha_{n,q} \cdot |x_{n,i,q} - x_{n,j,q}| \ ,
\end{equation}
where $\alpha_{n,q}$ are the weights participant $n$ implicitly assigns to issues $q$ for this aggregation.

Participant-level function $f_n$ is a linear or exponential kernel that translates the overall belief distance into a pixel distance, $d_{n,ij}$ (normalised to $0$ and $1$), that is, $f_n:[0,1]\mapsto [0,1]$. The value of $f_n$ increases monotonically with overall belief distance. We consider two kernel shapes (see Figure \ref{fig:kernels}): 

\begin{minipage}[t]{0.48\linewidth}
\begin{mdframed}
{\centering \textbf{Linear kernel function}}
\begin{equation}\label{eq:linear}
    f_n (x) = \hat a_n +  \hat b_n \cdot x  \ ,
\end{equation}
with $x$ the overall belief distance, i.e., the output of $g_n$, $a_n$ the minimum distance---i.e., how far apart individuals with zero overall belief distance are placed---, and $b_n$ the rate at which pixel distance increases with overall belief distance. 
\end{mdframed}
\end{minipage}
\hfill 
\begin{minipage}[t]{0.48\linewidth}
\begin{mdframed} 
{\centering \textbf{Exponential kernel function}} 
\begin{equation}\label{eq:expkernel}
    f_n (x') = 1 - \exp{\left(-\hat b_n \cdot (x') ^{\hat c_n}\right)}
\end{equation}
with similarity sharpness, $\hat{c}_n$---how abruptly similarity falls off with belief differences---, and the reach, $\hat{b}_n$---how far similarity extends across belief differences. Exponential kernels are a common choice for psychological similarity perception \citep[][]{roadsModelingSimilarityPsychological2024a}.
\end{mdframed}
\end{minipage}

\begin{figure}
    \centering
    \includegraphics[scale=1]{figures/kernels.png}
    \caption{Exponential and linear kernel functions $f_n$ translating overall belief distance, i.e., the output of $g_n$ into map distance $d_{n,ij}$.}
    \label{fig:kernels}
\end{figure}

We model map distances (equation \ref{eq:distance}) by fitting the parameters for each participant $n$---the sharpness $\hat{c}_n$, the reach $\hat{b}_n$ in equation \ref{eq:expkernel} (or the minimum distance $a_n$ and rate of distance increase $b_n$ for a linear kernel in equation \ref{eq:linear}), and the six (normalised) weights $\alpha_{n,q}$ (equation \ref{eq:weighted}). 


# Get issue weights


```{r}
# -------------------------------------------------------
# Helper: weighted L1 norm (g_n)
# alpha: length-6 vector (will be normalised to sum to 1)
# deltas: matrix of |x_{i,q} - x_{j,q}|, rows = pairs
# -------------------------------------------------------
weighted_l1 <- function(deltas, alpha) {
  # alpha_norm <- alpha / sum(alpha)          # normalise  # already ensured!
  as.vector(deltas %*% alpha)          # overall belief distance
}

# -------------------------------------------------------
# Kernel functions (f_n)
# -------------------------------------------------------
linear_kernel <- function(x, a, b) {
  a + b * x
}

exp_kernel <- function(x, a, b) {
  1 - exp(-b * (x ^ a))
}

# -------------------------------------------------------
# Loss function (SSE) — generic
# -------------------------------------------------------
make_loss <- function(deltas, pixel_dist, lambda, kernel = c("linear", "exponential")) {
  kernel <- match.arg(kernel)
  
  function(params) {
    # params layout:
    #   linear:      alpha[1:6], a, b        (8 params)
    #   exponential: alpha[1:6], a, b        (8 params)
    alpha  <- params[1:n_issues]
    g      <- weighted_l1(deltas, alpha)
    a <- params[n_issues + 1]  # intercept or exponent
    b <- params[n_issues + 2]  # slope/scaling of opinion difference
    
    if (kernel == "linear") {  
      pred <- linear_kernel(g, a, b)
    } else {
      pred <- exp_kernel(g, a, b)
    }
    
    sse <- sum((pixel_dist - pred)^2)
    entropy_penalty <- -lambda * sum(log(alpha + 1e-6))  # pulls weights from zero
    sse + entropy_penalty
  }
}

compute_diagnostics <- function(mat) {
  # VIF per column
  vifs <- setNames(numeric(ncol(mat)), colnames(mat))
  for (j in seq_len(ncol(mat))) {
    y <- mat[, j]
    X <- mat[, -j, drop = FALSE]
    if (nrow(X) <= ncol(X)) {
      vifs[j] <- NA_real_
    } else {
      r2 <- summary(lm(y ~ X))$r.squared
      vifs[j] <- 1 / (1 - r2)
    }
  }

  # Drop zero-variance columns before SVD
  col_sd <- apply(mat, 2, sd)
  mat_ok <- mat[, col_sd > 0, drop = FALSE]

  if (ncol(mat_ok) < 2) {
    # Can't compute a meaningful condition number with < 2 varying columns
    cond_number <- NA_real_
  } else {
    sv <- svd(scale(mat_ok))$d
    # Guard against near-zero singular values (numerical zero)
    min_sv <- min(sv[sv > .Machine$double.eps * max(sv)])
    cond_number <- max(sv) / min_sv
  }

  list(vif = vifs, condition_number = cond_number)
}

print("functions check")
```

```{r}

fit_participant <- function(df_p, kernel = c("linear", "exponential"), 
                            n_starts = 10, lambda = 0.0) {
  kernel <- match.arg(kernel)

  deltas     <- as.matrix(df_p[, delta_cols])
  pixel_dist <- df_p$pixel_dist
  loss       <- make_loss(deltas, pixel_dist, kernel, lambda = lambda)
  
  vif_vals <- compute_diagnostics(deltas)

  # if (kernel == "linear") {
  #   lb <- c(rep(0, n_issues), 0,    0)
  #   ub <- c(rep(1, n_issues), 1,    5)
  # } else {
  #   lb <- c(rep(0, n_issues), 1e-3, 1e-3)
  #   ub <- c(rep(1, n_issues), 50,   10)
  # }
  
  # heq <- function(params) sum(params[1:n_issues]) - 1
  
  # # --- run n_starts optimizations, keep best ---
  # best_fit    <- NULL
  # best_obj    <- Inf
  
  # for (i in seq_len(n_starts)) {
    
  #   # first start uses your original equal-weight x0, rest are random
  #   if (i == 1) {
  #     alpha0 <- rep(1 / n_issues, n_issues)
  #     if (kernel == "linear") {
  #       x0 <- c(alpha0, 0, 1)
  #     } else {
  #       x0 <- c(alpha0, 1, 4)
  #     }
  #   } else {
  #     raw    <- runif(n_issues)
  #     alpha0 <- raw / sum(raw)
  #     if (kernel == "linear") {
  #       x0 <- c(alpha0, runif(1, 0, 0.5), runif(1, 0.5, 2))
  #     } else {
  #       x0 <- c(alpha0, runif(1, 1, 10), runif(1, 0.5, 3))
  #     }
  #   }
    
  #   fit <- tryCatch(
  #     nloptr(
  #       x0          = x0,
  #       eval_f      = loss,
  #       lb          = lb,
  #       ub          = ub,
  #       eval_g_eq   = heq,
  #       opts        = list(
  #         algorithm = "NLOPT_LN_COBYLA",
  #         xtol_rel  = 1e-6,
  #         maxeval   = 10000
  #       )
  #     ),
  #     error = function(e) NULL
  #   )
    
  #   if (!is.null(fit) && fit$status > 0 && fit$objective < best_obj) {
  #     best_fit <- fit
  #     best_obj <- fit$objective
  #   }
  # }
  
  # # --- same result formatting as before ---
  # params     <- best_fit$solution
  # alpha_norm <- params[1:n_issues]
  
  # result <- tibble(
  #   issue      = delta_cols,
  #   alpha      = alpha_norm,
  #   vif        = vif_vals,
  #   kernel     = kernel,
  #   converged  = (best_fit$status > 0),
  #   sse        = best_fit$objective
  # )
  
  # if (kernel == "linear") {
  #   result$param1      <- params[n_issues + 1]
  #   result$param2      <- params[n_issues + 2]
  #   result$param1_name <- "linearKernel_param1"
  #   result$param2_name <- "linearKernel_param2"
  # } else {
  #   result$param1      <- params[n_issues + 1]
  #   result$param2      <- params[n_issues + 2]
  #   result$param1_name <- "expKernel_param1"
  #   result$param2_name <- "expKernel_param2"
  # }
  
  # result

  result <- tibble(
    issue      = delta_cols,
    # alpha      = alpha_norm,
    vif        = vif_vals$vif,
    condition_number = vif_vals$condition_number,
    # kernel     = kernel,
    # converged  = (best_fit$status > 0),
    # sse        = best_fit$objective
  )
  result
}
print("function fit one participant check")

```

```{r}
# -------------------------------------------------------
# Run for all participants × both kernels
# -------------------------------------------------------
set.seed(2)
results <- df_diff %>%
  group_by(id, wave) %>%
  group_modify(~ {

    res <- bind_rows(
      fit_participant(.x, kernel = "linear", lambda=0.0, n_starts = 1),
      fit_participant(.x, kernel = "exponential", lambda=0.0, n_starts = 1)
    )

    res

  }) %>%
  ungroup()
print(" ... fitting done.")
```

Analyse VIF:
```{r}
vif_summary <- results %>%
  group_by(id, wave) %>%
  summarise(
    max_vif  = max(vif, na.rm = TRUE),
    mean_vif = mean(vif, na.rm = TRUE),
    any_high = any(vif > 10, na.rm = TRUE),  # common threshold
    .groups  = "drop"
  )
```


Collect alphas:
```{r}
# -------------------------------------------------------
# Wide format: one row per participant × kernel
# -------------------------------------------------------
alphas_wide <- results %>%
  select(id, wave, kernel, issue, alpha, sse, converged, param1, param2,
         param1_name, param2_name) %>%
  pivot_wider(names_from = issue, values_from = alpha,
              names_prefix = "alpha_")

print(alphas_wide)

table(results$converged, results$kernel)
```

## Attach to dataframes

```{r}
alphas_join <- alphas_wide %>%

  # remove repeated metadata columns
  select(
    id, wave, kernel,
    starts_with("alpha_"),
    param1, param2,
    sse, converged
  ) %>%
  
  # rename params before pivoting
  mutate(
    param1 = case_when(
      kernel == "linear" ~ param1,
      kernel == "exponential" ~ param1
    ),
    
    param2 = case_when(
      kernel == "linear" ~ param2,
      kernel == "exponential" ~ param2
    )
  ) %>%
  
  pivot_wider(
    names_from = kernel,
    values_from = c(
      starts_with("alpha_"),
      param1,
      param2,
      sse,
      converged
    ),
    names_glue = "{kernel}_{.value}"
  )

names(alphas_join) <- names(alphas_join) %>%
  str_replace("alpha_deltaX_", "alpha_")


# select best kernel
alphas_join <- alphas_join %>%
  mutate(
    best_kernel = if_else(
      linear_sse < exponential_sse,
      "linear",
      "exponential"
    )
  )


```

Join back to df_diff
```{r}
df_diff_with_alphas <- df_diff %>%
  left_join(alphas_join,
            by = c("id", "wave"))
```

Join back to participant-level df_partic
```{r}
df_partic_with_alphas <- df_partic %>%
  left_join(alphas_join,
            by = c("id", "wave"))
```

```{r}
write.csv(df_partic_with_alphas, "processed_data/2026-05-19_data_processed_participant_withIssueWeights.csv", row.names=FALSE)

write.csv(df_diff_with_alphas, "processed_data/2026-05-19_data_processed_differences_withIssueWeights.csv", row.names=FALSE)
```




# Visualisation
THis is just some quick analysis of the inferred weights
## vis average 

```{r}

alpha_plot <- alphas_wide %>%
  pivot_longer(
    starts_with("alpha_"),
    names_to = "issue",
    values_to = "alpha"
  ) %>%
  mutate(
    issue = str_remove(issue, "alpha_")
  )

alpha_summary <- alpha_plot %>%
  group_by(kernel, issue) %>%
  summarise(
    mean = mean(alpha, na.rm = TRUE),
    se = sd(alpha, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
ggplot(alpha_summary,
       aes(x = issue,
           y = mean,
           fill = kernel)) +

  geom_col(position = "dodge") +

  geom_errorbar(
    aes(
      ymin = mean - 1.96 * se,
      ymax = mean + 1.96 * se
    ),
    position = position_dodge(.9),
    width = .2
  ) +

  coord_flip() +

  theme_minimal() +

  labs(
    title = "Average inferred issue weights",
    y = "Mean alpha"
  )
```


## Predict distances

```{r}


predict_distances <- function(df_p, fit_row) {
  alpha <- fit_row %>%
    select(starts_with("alpha_")) %>%
    as.numeric()
  deltas <- as.matrix(df_p[, delta_cols])
  g <- as.vector(deltas %*% alpha)
  if (fit_row$kernel == "linear") {
    pred <- fit_row$param1 +
      fit_row$param2 * g
  } else {
    pred <- 1 - exp(
      -fit_row$param1 * (g ^ fit_row$param2)
    )
  }
  tibble(
    observed = df_p$pixel_dist,
    predicted = pred
  )
}

participant = "330717073703941" # df_diff |> slice(1) |> pull(id)
w = 2
print(as.character(participant))
k = "linear"

one_fit <- alphas_wide %>%
  filter(kernel == k, id == participant, wave == w) 


one_df <- df_diff %>%
  filter(
    id == participant,
    wave == w
  )

pred_df <- predict_distances(one_df, one_fit)

ggplot(pred_df,
       aes(x = observed,
           y = predicted)) +

  geom_point(alpha = .6) +

  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "red"
  ) +

  coord_equal() +

  labs(
    title = paste("Observed vs predicted distances\n(participant ", participant, ", wave ", w, ", ", k, " kernel)"),
    x = "Observed pixel distance",
    y = paste("Predicted distance; ", k, " kernel")
  ) +

  theme_minimal()


```



# Analyse initial-condition dependence of weights

```{r}
# -------------------------------------------------------
# Multi-start fit for one participant
# -------------------------------------------------------
fit_participant_multistart <- function(df_p, kernel = c("linear", "exponential"), 
                                       n_starts = 50, lambda = 0) {
  kernel <- match.arg(kernel)
  
  deltas     <- as.matrix(df_p[, delta_cols])
  pixel_dist <- df_p$pixel_dist
  loss       <- make_loss(deltas, pixel_dist, kernel, lambda = lambda)
  
  if (kernel == "linear") {
    lb <- c(rep(0, n_issues), 0,    0)
    ub <- c(rep(1, n_issues), 1,    5)
  } else {
    lb <- c(rep(0, n_issues), 1e-3, 1e-3)
    ub <- c(rep(1, n_issues), 50,   10)
  }
  
  heq <- function(params) sum(params[1:n_issues]) - 1
  
  # --- run n_starts optimizations from random starting points ---
  runs <- map(seq_len(n_starts), function(i) {
    
    # random alpha0 on the simplex
    raw    <- runif(n_issues)
    alpha0 <- raw / sum(raw)
    
    if (kernel == "linear") {
      x0 <- c(alpha0, runif(1, 0, 0.5), runif(1, 0.5, 2))
    } else {
      x0 <- c(alpha0, runif(1, 1, 10), runif(1, 0.5, 3))
    }
    
    fit <- tryCatch(
      nloptr(
        x0          = x0,
        eval_f      = loss,
        lb          = lb,
        ub          = ub,
        eval_g_eq   = heq,
        opts        = list(
          algorithm = "NLOPT_LN_COBYLA",
          xtol_rel  = 1e-6,
          maxeval   = 10000
        )
      ),
      error = function(e) NULL
    )
    
    if (is.null(fit) || fit$status <= 0) return(NULL)
    
    params <- fit$solution
    tibble(
      start       = i,
      issue       = delta_cols,
      alpha       = params[1:n_issues],
      param1      = params[n_issues + 1],
      param2      = params[n_issues + 2],
      sse         = fit$objective,
      converged   = fit$status > 0
    )
  }) %>% 
    compact() %>%   # drop NULLs
    bind_rows()
  
  runs
}

# -------------------------------------------------------
# Summarise variance across starts for one participant
# -------------------------------------------------------
summarise_multistart <- function(runs) {
  runs %>%
    group_by(issue) %>%
    summarise(
      alpha_mean   = mean(alpha),
      alpha_sd     = sd(alpha),
      alpha_min    = min(alpha),
      alpha_max    = max(alpha),
      alpha_median = median(alpha),
      n_zero       = sum(alpha < 0.01),   # near-zero count
      n_starts     = n(),
      .groups = "drop"
    ) %>%
    mutate(
      is_unstable = alpha_sd > 0.1   # flag high-variance issues
    )
}

# -------------------------------------------------------
# Run on a single participant to inspect
# -------------------------------------------------------
test_id   <- 330183866219536   # replace with your participant
test_wave <- 1

runs_test <- df_diff %>%
  filter(id == test_id, wave == test_wave) %>%
  fit_participant_multistart(kernel = "linear", n_starts = 50, lambda = 0.1)

# summary table
summarise_multistart(runs_test)

# alpha distribution per issue across starts
runs_test %>%
  mutate(issue = factor(issue, levels = delta_cols)) %>%
  ggplot(aes(x = alpha)) +
  geom_histogram(bins = 20) +
  facet_wrap(~issue) +
  labs(title = paste("Alpha distributions across starts | id", test_id),
       x = "alpha", y = "count")

```
```{r}
# sse distribution — are there multiple basins?
runs_test %>%
  distinct(start, sse) %>%
  ggplot(aes(x = sse)) +
  geom_histogram(bins = 30) +
  labs(title = "SSE across starts", x = "SSE", y = "count")
```


```{r}

```